Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-17 doc/ApTranscriptome_manuscript.Rmd
completed draft results section

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of sympatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

@@

### Transcriptome assembly

@@

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the *Low* category (Fig. 1). For *A. carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A. picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* transcripts. Clustering of terms by semantic similarity yielded 3 clusters, with "regulation of biosynthetic/metabolic process" being the largest group.

For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 6 clusters, including "protein metabolic process", "lipid metabolic process", and "".

including some implicated in stress response (e.g. `r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][18]`) and cold tolerance (e.g. `r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][6]`) (Table S1). For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g. regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).

In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Figure 2). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).

@@

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock. For *High* genes, there were 4 enriched GO terms in *A. picea* and 12 in *A. carolinensis*. For *Low* genes, there were 40 enriched GO terms in *A. picea* and 23 in *A. carolinensis*. For *Intermediate* genes, there were 3 enriched GO terms in *A. picea* and 54 in *A. carolinensis*. For *Bimodal* genes, there were 12 enriched GO terms in *A. picea* and 10 in *A. carolinensis*.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively. We found that nearly 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress. The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A. carolinensis* and more up-regulation of gene expression in *A. picea* in response to thermal stress (Fig. 2).

>>>>

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

@@

@@

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. Thus, at a minimum ... of genes are thermally-responsive. However, the closest related ant species with a sequenced genome has only ... genes. If we use this as

The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the *Low* category (Fig. 1). For *A. carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A. picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For *Intermediate* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further performed GSEA for the genes that shifted from being shut-down at extreme temperatures, *Intermediate*, in *A. carolinensis* to having increased expression in *Low* temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For *Intermediate* genes in *A. carolinensis* that shifted *Bimodal* in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

### Candidate genes

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`).

@@

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively. We found that at least 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress. The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A. carolinensis* and more up-regulation of gene expression in *A. picea* in response to thermal stress (Fig. 2).
